## RSTUDIO

#### 4 panels
Top left: Script viewer - write your code to save it for future use
Bottom left: Console - this is where R is actually running
Top right: Variable/history displayer
Bottom right: Files/plots/packages/help displayer

#### Script vs Console
Console: type commands directly to see their result. This is not saved
Script editor: Save your code for later. 

#### Shortcuts
TAB -- autocomplete
control/cmd + enter -- run current line from the script in console

## How to run a line of code?
When you are on the line, press cmd (mac) / ctrl (windows), then enter (more later)

## Comments

Comments are lines in the script that are skipped and not run. They can be used to annotate your code and make it more friendly when you are trying out something new or for others if you are sharing your code. This is also VERY helpful to remind yourself of what you did when you look back at the code years after you wrote it.

If you add a hash, everything written after is a comment.

```{r comments2, error=TRUE}

# This is a comment. We can use comments using a hash sign
This is not a comment

```




## Setting Working Directory

directory = folder

The folder R will load files from and save files to. You can check and change the current directory to a different one

```{r directory2}

# Where am I? i.e. where is R currently looking for files or saving files to?
getwd()

# Set the location i.e.working directory.  This is where R will look for files you want to load.
# setwd()

# sample working directory for MAC users
# setwd("/Users/hoang/Desktop")

# sample working directory for windows users
setwd("C:/Users/hoang/Documents/demo")
getwd()
#if you get this warning 
#Warning: The working directory was changed to C:/Users/hoang/Documents/demo inside a notebook chunk. The working directory will be reset when the chunk is finished running. Use the knitr root.dir option in the setup chunk to change the working directory for notebook chunks. Then use the copy the code below and place it at the beginning of this r markdown (.rmd) file
```

```{r setup, include=FALSE}
#knitr::opts_knit\$set(root.dir = "C:/Users/hoang/Documents/demo")
```




## Getting Help in RStudio

Below are some ways to get help within RStudio. You can, of course, google functions and specific things you are looking for. R Documentation is usually a good first source because it details everything a function can do, but if you are looking for something more specific, often someone else has asked the same question on public forums like StackOverflow

```{r help2, include=FALSE}

help.start()   # general help

help(mean)    # help for function mean()
?mean         # same as above

example(mean) # shows an example of function mean()
```



## Packages

Packages in R (and really any programming language) allow you to add functionality to base R. So you can are downloading a package of functions that did not exist before in base R.

For example, if you want to do mixed models in R, you would download a package called 'lme4'. You can't do that with base R. We also download packages which make our lives easier and code easier to read and understand. For example, we can create plots to visualize data in base R but they are pretty tedious if you want to make them pretty and publication-ready, you would probably use 'ggplot2'. 

There are alot of packages that add functionality or make our code easier to read. Hadley Wickham (http://hadley.nz/) made a universe of useful packages so you don't have to install each one of them individually. 'tidyverse' is a universe of such useful packages. When you load 'tidyverse', you will get a huge print out in your console detailing the packages that come with tidyverse

```{r packages2}

## Installing Packages
# uncomment the line below and run it if you don't have this package installed
#install.packages('tidyverse')

## to view a list of installed packages
# installed.packages()

## Loading Packages
# If you haven't installed the 'tidyverse' package, you should get an error when you run the line below
# If you have it installed, you will get a huge print out in your console detailing the packages that come with tidyverse

library(tidyverse)
```




# R Resources

Free online edX course by Rafael Irizarry, Harvard University: https://www.edx.org/course/data-science-r-basics 

**Cheatsheets**
  *note that some of these are links to PDFs which will download when you click on them

https://rstudio.com/resources/cheatsheets/ (various cheatsheets)
Tidyverse cheatsheet: https://www.datacamp.com/community/blog/tidyverse-cheat-sheet-beginners
RStudio cheatsheet: https://rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf 
Base R cheatsheet: https://rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf


```{r, }
#Simple arithmatic operations
2
5+4
3*3
10/2

#A variable is a container that holds information like values or words
a <- 5 # variable (numeric)
class(a)

b <- 7 + 3.0


c <- a+b
c <- a-b
c <- a*b
c <- a/b 


d <- 'Hello World' # character variable
class(d)
d[1]

d <- '3'
class(d)

#Vector
#it is the most basic data structure in R. It is a sequence of data elements of the exact same type (such as all numbers or all text strings).

# A numeric vector
grades <- c(40, 55, 77, 80, 89, 99)

# A character (text) vector
cities <- c("Toronto", "New York", "London")

# A logical (boolean) vector
answers <- c(TRUE, FALSE, TRUE)

#clear these values with broom icon 
#then re-run codes via RUN function(green play button at top right corner of this grey box)
#find function
#highlight variable to where else it appears in your markdown
#pressing tab on the keyboard to quickly complete your variable of interest
#pressing up arrow in the console will quickly highlight your last function used
```
```{r,}
## create a vector (R likes to call them lists) by using the c() function called combine or concatenate
list <- c(44,21,22,30,50,84) # a variable called 'list' of six numbers
r <- c(44,21,22,30,50,84)
call_it_whatever_you_like <- c(44,21,22,30,50,84)


## create a range of values
list2 <- seq(1,10)
list2b <- 1:10 # : works too, but no step option
list3 <- seq(1,10,by=2) # seq includes a step size with 'by' argument

## index into a vector
list[1]
list[3]
list[1:3] # : range of indices
list[c(1,3,4)] # index with vector
list[c(1:3,6)] # can include ranges and single indices
list < 30 # standard logical operations 
list[list<30] # ...can be used for indexing

#what about characters like my variable d of "hello world"?
library(stringr)
d <- 'hello world boo!'
# Directly extracts the 3rd word
str_split_i(d, " ", 3)

## vector calculations (elementwise)
list4 <- list*10
list4 <- list[1:2]*10
list4 <- list - list2
list5 <- (list+list2)/2

view(list5)
class(list5)

## create matrix and index it
m <- matrix(21:39,nrow=3,ncol=4)
m
m[3] # single index points to the specific row position of the first column
m[2,3] # row and column indexing
m[,4] # 3rd column (nothing can mean something in R!)
m[2,] # 2nd row

```


```{r, }

#Replicate function
#The rep() function in R is a base utility used to replicate elements of vectors and lists. It allows you to create larger data structures by repeating existing values in specific patterns

#Example1 with times: Repeats the entire vector a specific number of times.
a <- rep(1:2, times = 3) #→ 1, 2, 1, 2, 1, 2


#Example2 with each: Repeats each individual element of the vector before moving to the next.
b <- rep(1:2, each = 3) #→ 1, 1, 1, 2, 2, 2

stress_exercise <- c(2,2,4,5,5)
view(stress_exercise)

stress_NOexercise <- c(11,15,20,13,22)

#create data frame with both exercise and no exercise
group <- rep(c("exercise","no_exercise"), each=5)
view(group)  
class(group)

stress_data <- data.frame (
  group = rep(c("exercise","no_exercise"), each=5)
  
)

stress_data <- data.frame (
  group = rep(c("exercise","no_exercise"), each=5),
  stress = c(stress_exercise, stress_NOexercise)
)


#add a new column
stress_data$sleep <- c(7,8,7,6,8,9,7,6,7,8)
summary(stress_data)
```

```{r, }
#### --- T-TESTs AND GRAPH --- ####

#convert character to factor for analysis
stress_data$group <- as.factor(stress_data$group)

#do t-test for sleep
sleep.group <- t.test(sleep ~ group, data=stress_data)
sleep.group

#do t-test for stress
stress.group <- t.test(stress ~ group, data=stress_data)
stress.group

#plot
graph_stress <- ggboxplot(stress_data, x = "group", y = "stress",
          color = "group",
          ylab = "stress", xlab="group")

graph_stress

```

```{r, }

# --- Bootstrap to 100 participants (50 per group) ---
#Bootstrapping is a statistical method that resamples your original dataset with replacement to estimate the accuracy, confidence intervals, or standard errors of a statistic (like a mean, median, or regression coefficient) 
#In R, it means creating hundreds or thousands of "pseudo-datasets" of the exact same size as your original data, calculating your statistic on each, and analyzing the distribution of those results.

set.seed(123)  # for reproducibility
#ensures you get the same "random" results every time you run your code.


# Split by group and resample each separately
exercise_rows    <- stress_data[stress_data$group == "exercise", ]
no_exercise_rows <- stress_data[stress_data$group == "no_exercise", ]

#The sample() function in R is used to take a random sample of a specified size from a dataset or vector. It is a core tool for tasks like data analysis, simulations, and randomized experiments
boot_exercise <- exercise_rows[
  sample(nrow(exercise_rows), size = 50, replace = TRUE), ]
#The sample(nrow(df)) expression is a specific coding technique used to randomly shuffle the rows of a data frame. It generates a completely randomized list of row indices.
#replace: If TRUE, an item can be picked more than once (sampling with replacement). If FALSE (default), each item is picked only once.

boot_no_exercise <- no_exercise_rows[
  sample(nrow(no_exercise_rows), size = 50, replace = TRUE), ]

# Combine and reset row names
stress_data_100 <- rbind(boot_exercise, boot_no_exercise)
rownames(stress_data_100) <- 1:100

# Inspect result
summary(stress_data_100)
table(stress_data_100$group)   # should show 50 / 50
head(stress_data_100, 25)

#convert character to factor for analysis
stress_data_100$group <- as.factor(stress_data_100$group)

#do t-test for sleep
sleep.group.100 <- t.test(sleep ~ group, data=stress_data_100)
sleep.group.100

#do t-test for stress
stress.group.100 <- t.test(stress ~ group, data=stress_data_100)
stress.group.100

#plot
graph_stress_100 <- ggboxplot(stress_data_100, x = "group", y = "stress",
          color = "group",
          ylab = "stress", xlab="group")

graph_stress_100
```




```{r, }
# Boolean = Logical data
# a boolean is a data type that represents logical truth values, but it is officially called a logical data type. It can only hold one of two values: TRUE or FALSE.

z <- TRUE 
class(z)
y <- T # logical shorthands!
x <- F
w <- F

## logical operators
z & y # and operator
z & x
z | y # or operator
z | w

3 == 4 # equivalence
3 != 4 # non equivalence
3 < 4
(5 > 4) & (8 > 10)
(2 < 4) & (5 < 10)
(5 > 4) | (8 > 10)
(5 < 4) | (8 > 10)

# Sample vector of ages
ages <- c(15, 25, 40, 65)

# Example 1: Age is greater than 18 AND less than 60
ages > 18 & ages < 60

# Example 2: Age is under 18 OR over 60
ages < 18 | ages >= 60

# Example 3
user_role <- "admin"
logged_in <- FALSE

# R will waste time checking the user role, even if logged_in is FALSE which will return an error instead of letting the code bypass it. 
#if (logged_in & user_role2 == "admin") {
  #print("Access granted to dashboard.")
#}

# With '&&' R won't waste time checking the user role; this is why '&&" is called short-circuit
if (logged_in && user_role2 == "admin") {
  print("Access granted to dashboard.")
}
#Always use && and || inside if (...) statements in R. 
#Use & and | exclusively for filtering columns or vectors of data.
```

```{r, }
##### --- CONDITIONAL STATEMENTS --- #####
#Conditional statements are code structures used to make decisions in R. They execute specific blocks of code only if a preset logical condition evaluates to TRUE

#1. The if Statement
#The most basic structure. It runs the code inside the curly braces {} only if the condition in the parentheses () is true. 

score <- 85

if (score >= 50) {
  print("Hurray, you passed!")
}

#2. The if-else Statement
#This provides an alternative path. If the condition is FALSE, the code inside the else block runs instead. 
score <- 42

if (score >= 50) {
  print("You passed!")
} else {
  print("Whop whop, you failed.")
}

#3. The if-else if-else Chain
#Used when you need to test multiple, distinct conditions in a specific order.
score <- 75

if (score >= 90) {
  print("Grade: A")
} else if (score >= 75) {
  print("Grade: B")
} else {
  print("Grade: C")
}

## basic control flow
## if/else commands
nice_weather <- T
very_sunny <- T
if(nice_weather==T && very_sunny==T) {
  print("let's go outside, put on sunscreen!")
} else if(very_sunny==T) {
  print("put on sunscreen!")
} else {
  print("more R inside!")
}


```














